---
title: "Combine PhysioNet Challenge 2019 Files"
output:
  html_document:
    toc: yes
    number_sections: yes
  html_notebook:
    toc: yes
---

<style type="text/css">
div#TOC li {
    list-style:none;
    background-image:none;
    background-repeat:none;
    background-position:0;
}
</style> 

efg | 2019-04-05

>An expanded training database is now available, containing data from a total of 40,336 subjects. The complete training database (42 MB), consisting of two parts: training set A (20,336 subjects) and B (20,000 subjects).

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
time.1 <- Sys.time()  
```

Combine all 2019 challenge files by directory into a single file for analysis.

# Setup

Directory structure

```
PhysioNet-2019
    raw
    data
        trainingA
          [20336 files]
        trainingB
          [20000 files]
    R
        000-Download
        010-Count-Characters
        020-Combine-Files
        ...
```

## Packages

```{r, comment=NA}
library(tidyverse)
library(kableExtra) # kable_styling
library(writexl)
```

## Helper function

```{r}
Show <- function(data, caption="", bigMark="")
{
  data                                          %>%
  kable("html",
        caption=caption,
        format.args = list(big.mark=bigMark))   %>%
  kable_styling(bootstrap_options = c("striped", "bordered", "condensed"),
                position = "left", full_width = FALSE)
}
```                  

# Make list of PhysioNet .psv files by directory

Assume files previously unzipped into this directory relative to project directory.

```{r, comment=NA}
dataDirA <- "../../data/trainingA" 
filesA <- list.files(dataDirA, full.names=TRUE)
length(filesA)
```

```{r, comment=NA}
dataDirB <- "../../data/trainingB" 
filesB <- list.files(dataDirB, full.names=TRUE)
length(filesB)
```

# Combine all files into single tibble

Add `Source` of `A` or `B` to track origin of data

## Function to read single .psv file

Add filename as first column.  Move ICULOS to second column.

Convert NaNs to NAs while reading so tidyr fill function can be used more easily later.

```{r}
readPSV <- function(filepath)
{
  filename <- substr(basename(filepath), 1, nchar(basename(filepath))-4)
  d <-  
    read_delim(filepath, "|", 
               col_types = cols(.default = col_double()),
               na = "NaN")                                    %>%
    mutate(patient = filename)                                %>%
    select(patient, ICULOS, everything())
  
  d
}
```

## map_df from purrr does heavy lifting

[Reading and combining many tidy data files in R](https://serialmentor.com/blog/2016/6/13/reading-and-combining-many-tidy-data-files-in-R).

```{r, comment=NA}
allDataA <- 
  filesA               %>%
  map_dfr(readPSV)     %>% 
  mutate(Source = "A") %>%
  select(Source, everything())

nrow(allDataA)
```

```{r, comment=NA}
allDataB <- 
  filesB               %>%
  map_dfr(readPSV)     %>% 
  mutate(Source = "B") %>%
  select(Source, everything())

nrow(allDataB)
```

## Structure of tibbles

Should be the same

### A

```{r, comment=NA}
str(allDataA)
```

### B

```{r, comment=NA}
str(allDataB)
```

## Write to Excel file

```{r}
write_xlsx(allDataA, "../../data/PhysioNet-2019-Raw-Data-A.xlsx")
write_xlsx(allDataB, "../../data/PhysioNet-2019-Raw-Data-B.xlsx")
```

## Save R object for quicker loading later

```{r}
save(allDataA, file = "../../data/PhysioNet-2019-Raw-Data-A.RData")
save(allDataB, file = "../../data/PhysioNet-2019-Raw-Data-B.RData")
```

# Record counts

## A

```{r}
allDataCountsA <- 
  allDataA                                     %>%
  summarize(nRecord  = n(),
            nPatient = n_distinct(patient))

allDataCountsA %>%
  Show(bigMark = ",")
```

## B

```{r}
allDataCountsB <- 
  allDataB                                     %>%
  summarize(nRecord  = n(),
            nPatient = n_distinct(patient))

allDataCountsB %>%
  Show(bigMark = ",")
```

## Sepsis records

```{r}
sepsisDataCountsA <-
  allDataA                   %>%
  filter(SepsisLabel == 1)   %>%
  summarize(nRecord  = n(),
            nPatient = n_distinct(patient)) 

sepsisDataCountsA %>%
  Show(bigMark = ",")
```

```{r}
sepsisDataCountsB <-
  allDataB                   %>%
  filter(SepsisLabel == 1)   %>%
  summarize(nRecord  = n(),
            nPatient = n_distinct(patient)) 

sepsisDataCountsB %>%
  Show(bigMark = ",")
```

## Percent Sepsis

### A

```{r, comment=NA}
sepsisDataCountsA$nRecord
allDataCountsA$nRecord

sprintf("%.2f%% sepsis records",
        100 * sepsisDataCountsA$nRecord  / allDataCountsA$nRecord)
```

```{r, comment=NA}
sepsisDataCountsA$nPatient
allDataCountsA$nPatient

sprintf("%.2f%% sepsis patients",
        100 * sepsisDataCountsA$nPatient / allDataCountsA$nPatient)
```

### B

```{r, comment=NA}
sepsisDataCountsB$nRecord
allDataCountsB$nRecord

sprintf("%.2f%% sepsis records",
        100 * sepsisDataCountsB$nRecord  / allDataCountsB$nRecord)
```

```{r, comment=NA}
sepsisDataCountsB$nPatient
allDataCountsB$nPatient

sprintf("%.2f%% sepsis patients",
        100 * sepsisDataCountsB$nPatient / allDataCountsB$nPatient)
```

# Sample records

```{r}
head(allDataA)  %>% 
  Show()        %>%
  scroll_box(width = "1000px")
```

```{r}
tail(allDataB) %>% 
  Show()       %>%
  scroll_box(width = "1000px")
```

```{r, echo=FALSE}
time.2 <- Sys.time()
processingTime <- paste("Processing time:", sprintf("%.1f",
                        as.numeric(difftime(time.2,
                                            time.1, units="secs"))), "secs\n")
```

`r processingTime`
`r format(time.2, "%Y-%m-%d %H%M")`                      