---
title: "Download Physionet 2019 Challenge"
output:
  html_document:
    toc: yes
    number_sections: yes
  html_notebook:
    toc: yes
---

<style type="text/css">
div#TOC li {
    list-style:none;
    background-image:none;
    background-repeat:none;
    background-position:0;
}
</style>
                           
efg | 2019-04-01

>An expanded training database is now available, containing data from a total of 40,336 subjects. The complete training database (42 MB), consisting of two parts: training set A (20,336 subjects) and B (20,000 subjects).

Start at [Physionet 2019 Challenge page](https://physionet.org/challenge/2019/) for general information.

Here's the [download page](https://physionet.org/users/shared/challenge-2019/).

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
time.1 <- Sys.time()
```

# Setup

```{r}
suppressPackageStartupMessages( library(R.utils) )   # downloadFile, gunzip
suppressPackageStartupMessages( library(dplyr) )
```

```{r}
rawDIR  <- "../../raw/"
dataDIR <- "../../data" 
```

# Directories

A *raw* directory will be created for the downloadeded ZIP, which will be unzipped to form the *data* directory.

The *raw* directory could be deleted at the conclusion of this script, but that is left as a manual operation.

```{r}
if (!dir.exists(rawDIR)) dir.create(rawDIR)
if (!dir.exists(dataDIR)) dir.create(dataDIR)
```

# Download ZIP file 

```{r, comment=NA}
URL <- "https://physionet.org/users/shared/challenge-2019/"
```

```{r, comment=NA}
filenameA <- "training_setA.zip"
downloadFile(paste0(URL, filenameA), path = rawDIR, mode = "wb",  quiet = TRUE)
```

```{r, comment=NA}
filenameB <- "training_setB.zip"
downloadFile(paste0(URL, filenameB), path = rawDIR, mode = "wb",  quiet = TRUE)
```

# Unzip

```{r, comment=NA}
unzip(paste0(rawDIR, filenameA), exdir = "../../data", setTimes = TRUE)
```

```{r, comment=NA}
unzip(paste0(rawDIR, filenameB), exdir = "../../data", setTimes = TRUE)
```

```{r, echo=FALSE}
time.2 <- Sys.time()
processingTime <- paste("Processing time:", sprintf("%.1f",
                        as.numeric(difftime(time.2,
                                            time.1, units="secs"))), "secs\n")
```

`r processingTime`
`r format(time.2, "%Y-%m-%d %H%M")`
