---
title: "Laboratory Values Stats - PhysioNet Sepsis Challenge"
output:
  html_document:
    toc: yes
    number_sections: yes
  html_notebook:
    toc: yes
---

<style type="text/css">
div#TOC li {
    list-style:none;
    background-image:none;
    background-repeat:none;
    background-position:0;
}
</style>

efg | 2019-04-01

>An expanded training database is now available, containing data from a total of 40,336 subjects. The complete training database (42 MB), consisting of two parts: training set A (20,336 subjects) and B (20,000 subjects).

efg | 2019-05-12

Updated units for Platelets and WBC after PhysioNet forum posting on 2019-04-17 by Gari Clifford.

efg | 2019-05-19

Encountered [pandoc document conversion failed with error 11 #4462](https://github.com/rstudio/rstudio/issues/4462) after R upgrade to 3.6.0 and latest RStudio.  Fix in link from RStudio resolved the problem.
 
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
time.1 <- Sys.time()
```

# Purpose

The purpose of this notebook is to take a preliminary look at the raw data from 40,336 patients in the PhysioNet 2019 challenge, where some had sepsis and some did not.  

The focus of this notebook is on the **Laboratory Values** data (**columns 9 - 34**) and the outcome (column 41)

* `SepsisLabel`.  For sepsis patients, `SepsisLabel` is 1 if $t \ge t_{sepsis} - 6$ and 0 if $t < t_{sepsis} - 6$. For non-sepsis patients, `SepsisLabel` is 0. 

# Background

The data being analyzed is from the [Early Prediction of Sepsis from Clinical Data: the PhysioNet/Computing in Cardiology Challenge 2019](https://physionet.org/challenge/2019/).

# Exploration

## Setup

```{r}
plotCaptionLeft  <- "Source:  PhysioNet 2019 Sepsis Project"
plotCaptionRight <- paste("efg", format(Sys.time(), "%Y-%m-%d"))
```

### Packages

```{r, comment=NA}
library(tidyverse)
library(kableExtra) # kable_styling
library(readxl)

library(gridExtra)
library(grid)

library(e1071)      # skewness, kurtosis
```

### Helper function

```{r}
Show <- function(data, caption="", bigMark="")
{
  data                                          %>%
  kable("html",
        caption=caption,
        format.args = list(big.mark=bigMark))   %>%
  kable_styling(bootstrap_options = c("striped", "bordered", "condensed"),
                position = "left", full_width = FALSE)
}
```
            
### Laboratory Values List and Units

From file `Laboratory-Values.xlsx` but order by `Quantity` label.
           
```{r}
labValues <- 
  read_xlsx("Laboratory-Values.xlsx")              %>%
  mutate(Units = ifelse(is.na(Units), "", Units))  %>%
  arrange(Quantity)

labValues %>%
  Show()
```

The lab values are sorted here by Quantity name, which is a different order than in the file.

### Load PhysioNet 2019 Challenge Combined Data

```{r, comment=NA}
load(file = "../../data/PhysioNet-2019-Raw-Data-A.RData")
dim(allDataA)
```

```{r, comment=NA}
load(file = "../../data/PhysioNet-2019-Raw-Data-B.RData")
dim(allDataB)
```

Combine A and B

```{r}
allData <- bind_rows(allDataA, allDataB)
dim(allData)
```

Verify Laboratory Values match

```{r, comment=NA}
all(labValues$Quantity %in% names(allData))
```

Sample Laboratory Values

```{r}
allData                        %>%
  select( patient, ICULOS, 
          labValues$Quantity,
          SepsisLabel)         %>%
  head()                       %>%
  Show()                       %>%
  scroll_box(width = "800px")
```

## Sepsis Label

```{r, comment=NA}
table(allData$SepsisLabel, useNA = "ifany")
```

## Laboratory Values

```{r, results="asis", fig.width=8, fig.height=6}
for (i in 1:nrow(labValues))
{
  titleLine <- labValues$Quantity[i]
  if (titleLine != labValues$Description[i])
  {
    titleLine <- paste(titleLine, " - ", labValues$Description[i])
  }
  
  cat("\n###", titleLine, "\n")   # dynamic header in markdown
  
  if (labValues$Quantity[i] %in% c("Platelets", "WBC"))
  {
    cat("\nUnits updated after PhysioNet forum posting by Gari Clifford on 2019-04-17\n")
  }               
  
  rawData <- 
    allData                                      %>%
    select(labValues$Quantity[i], SepsisLabel)   %>%
    na.omit()                                    %>%
    mutate(Sepsis = recode(SepsisLabel,
                           "0" = "Not Sepsis",
                           "1" = "Sepsis"))      %>%
    mutate(Sepsis = as.factor(Sepsis))
  
  counts <- table(rawData$Sepsis)
  rawDataLabels <- names(counts)
  rawDataCounts <- str_trim( format(counts, big.mark = ",") )
  
  if (nchar(labValues$Units[i]) == 0)
  {
    quantityUnits <- ""    # display no units
  } else {
    quantityUnits <- paste0("[", labValues$Units[i], "]")
  }
   
  valueBoxplot <- 
    ggplot(rawData, 
           aes_string("Sepsis", 
                      labValues$Quantity[i], 
                      fill = "Sepsis"))                        +
    geom_boxplot(notch = TRUE)                                 +
    ylim(labValues$MinPlotValue[i], labValues$MaxPlotValue[i]) +
    scale_x_discrete(labels = paste(rawDataLabels,
                                    paste0("n = ", 
                                           rawDataCounts), 
                                    sep = "\n") )              +
    labs(title = titleLine,
         x = "",
         y = paste0(labValues$Quantity[i], quantityUnits))     +
    scale_fill_brewer(palette = "Pastel1", direction = -1)     +
    theme(legend.position = "none",
          plot.margin = unit(c(5, 5, 5, 12), "mm"))            +
    coord_flip()
  
  valueDensity <- 
    ggplot(rawData, aes_string(x = labValues$Quantity[i]))     +
    geom_density(aes_string(group = "Sepsis", fill = "Sepsis"), 
                 alpha = 0.5)                                  +
    xlim(labValues$MinPlotValue[i], labValues$MaxPlotValue[i]) +
    labs(x = paste0(labValues$Quantity[i], quantityUnits),
         caption = c(plotCaptionLeft, plotCaptionRight))       +
    scale_fill_brewer(palette = "Pastel1", direction = -1)     +
    theme(plot.caption = element_text(hjust = c(0, 1)),
          legend.position = "none",
          plot.margin = unit(c(5, 5, 5, 20), "mm"))
  
  suppressWarnings( grid.arrange(valueBoxplot, valueDensity) )
  
  # mulitple quantile calls not efficient, but fast enough here
  stats <-
    rawData           %>%
    group_by(Sepsis)  %>%
    summarize(N        = n(),
              Distinct = n_distinct(.[1]),
              Mean     = mean(!!rlang::sym(names(rawData)[1])),
              StdDev   = sd(!!rlang::sym(names(rawData)[1])),
              Skewness = skewness(!!rlang::sym(names(rawData)[1])),
              Kurtosis = kurtosis(!!rlang::sym(names(rawData)[1])),
              Min      = min(!!rlang::sym(names(rawData)[1])),
              Q5       = quantile(!!rlang::sym(names(rawData)[1]), 0.05),
              Q25      = quantile(!!rlang::sym(names(rawData)[1]), 0.25),
              Median   = median(!!rlang::sym(names(rawData)[1])),
              Q75      = quantile(!!rlang::sym(names(rawData)[1]), 0.75),
              Q95      = quantile(!!rlang::sym(names(rawData)[1]), 0.95),
              Max      = max(!!rlang::sym(names(rawData)[1])))   %>%
    ungroup()         %>%
    arrange(desc(Sepsis))
  
  print( stats %>%  Show(bigMark = ",") )
  
  cat("\n")
}
```


Notes:

* StackOverflow [R dplyr operate on a column known only by its string name](https://stackoverflow.com/questions/46310123/r-dplyr-operate-on-a-column-known-only-by-its-string-name).

* StackOverflow [Using dplyr window functions to calculate percentiles](https://stackoverflow.com/questions/30488389/using-dplyr-window-functions-to-calculate-percentiles).

# Discussion

Need to impose limits since outliers distort some charts too much.

[work in progress]

# Conclusions

[work in progress]

# sessionInfo

```{r}
library(rmarkdown)
pandoc_version()
```

```{r}
sessionInfo()
```

```{r, echo=FALSE}
time.2 <- Sys.time()
processingTime <- paste("Processing time:", sprintf("%.1f",
                        as.numeric(difftime(time.2,
                                            time.1, units="secs"))), "secs\n")
```

`r processingTime`
`r format(time.2, "%Y-%m-%d %H%M")`                                     