---
title: "Vital Signs Stats - PhysioNet Sepsis Challenge"
output:
  html_document:
    toc: yes
    number_sections: yes
  html_notebook:
    toc: yes
---

<style type="text/css">
div#TOC li {
    list-style:none;
    background-image:none;
    background-repeat:none;
    background-position:0;
}
</style>

efg | 2019-03-03

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
time.1 <- Sys.time()
```

# Purpose

The purpose of this notebook is to take a preliminary look at the raw data from 5000 patients in the PhysioNet 2019 challenge, where some had sepsis and some did not.  

The focus of this notebook is on the Vital Signs data (columns 1 - 8) and the outcome (column 41)

* `SepsisLabel`.  For sepsis patients, `SepsisLabel` is 1 if $t \ge t_{sepsis} - 6$ and 0 if $t < t_{sepsis} - 6$. For non-sepsis patients, `SepsisLabel` is 0. 

# Background

The data being analyzed is from the [Early Prediction of Sepsis from Clinical Data: the PhysioNet/Computing in Cardiology Challenge 2019](https://physionet.org/challenge/2019/).

# Exploration

## Setup

```{r}
plotCaptionLeft  <- "Source:  PhysioNet 2019 Sepsis Project"
plotCaptionRight <- paste("efg", format(Sys.time(), "%Y-%m-%d"))
```

### Packages

```{r, comment=NA}
library(tidyverse)
library(kableExtra) # kable_styling
library(readxl)

library(gridExtra)
library(grid)
```

### Helper function

```{r}
Show <- function(data, caption="", bigMark="")
{
  data                                          %>%
  kable("html",
        caption=caption,
        format.args = list(big.mark=bigMark))   %>%
  kable_styling(bootstrap_options = c("striped", "bordered", "condensed"),
                position = "left", full_width = FALSE)
}
```
            
### Vital Signs List and Units
           
```{r, echo=FALSE}
vitalSigns <- 
  read_xlsx("Vital-Signs.xlsx") 

vitalSigns %>%
  Show()
```

### Load PhysioNet 2019 Challenge Combined Data

```{r, comment=NA}
load(file = "../../data/PhysioNet-2019-Raw-Data.RData")
dim(allData)
```

Verify vital signs match

```{r, comment=NA}
all(vitalSigns$Quantity %in% names(allData))
```

Sample Vital Signs

```{r}
allData                                           %>%
  select( patient, ICULOS, 
          vitalSigns$Quantity,
          SepsisLabel)                            %>%
  head()                                          %>%
  Show()  
```

## Sepsis Label

```{r, comment=NA}
table(allData$SepsisLabel, useNA = "ifany")
```

## Vital Signs

```{r, results="asis", fig.width=8, fig.height=6}
for (i in 1:nrow(vitalSigns))
{
  cat("\n###", vitalSigns$Description[i], "\n")
  
  rawData <- 
    allData                                      %>%
    select(vitalSigns$Quantity[i], SepsisLabel)  %>%
    na.omit()                                    %>%
    mutate(Sepsis = recode(SepsisLabel,
                           "0" = "Not Sepsis",
                           "1" = "Sepsis"))      %>%
    mutate(Sepsis = as.factor(Sepsis))
   
  valueBoxplot <- 
    ggplot(rawData, 
           aes_string("Sepsis", 
                      vitalSigns$Quantity[i], 
                      fill = "Sepsis"))                      +
    geom_boxplot(notch = TRUE)                               +
    scale_x_discrete(labels = paste(c("Not Sepsis", "Sepsis"),
                                    paste0("n = ", 
                                           table(rawData$Sepsis)) , 
                                    sep = "\n") ) +
    labs(title = vitalSigns$Description[i],
         x = "",
         y = paste0(vitalSigns$Description[i], 
                          "[", vitalSigns$Units[i], "]"))    +
    scale_fill_brewer(palette = "Pastel1", direction = -1)   +
    theme(legend.position = "none",
          plot.margin = unit(c(5, 5, 5, 12), "mm"))          +
    coord_flip()
  
  valueDensity <- 
    ggplot(rawData, aes_string(x = vitalSigns$Quantity[i]))  +
    geom_density(aes_string(group = "Sepsis", fill = "Sepsis"), 
                 alpha = 0.5)   +
    labs(title = vitalSigns$Description[i],
         x = paste0(vitalSigns$Description[i], 
                          "[", vitalSigns$Units[i], "]"),
         caption = c(plotCaptionLeft, plotCaptionRight))     +
    scale_fill_brewer(palette = "Pastel1", direction = -1)   +
    theme(plot.caption = element_text(hjust = c(0, 1)),
          legend.position = "none",
          plot.margin = unit(c(5, 5, 5, 20), "mm"))
  
  grid.arrange(valueBoxplot, valueDensity)
  
  cat("\n")
}
```

# Discussion

[work in progress]

# Conclusions

[work in progress]

```{r, echo=FALSE}
time.2 <- Sys.time()
processingTime <- paste("Processing time:", sprintf("%.1f",
                        as.numeric(difftime(time.2,
                                            time.1, units="secs"))), "secs\n")
```

`r processingTime`
`r format(time.2, "%Y-%m-%d %H%M")`                                     