---
title: "Physionet Challenge Count Characters"
output:
  html_document:
    df_print: paged
---

Create frequency counts of all characters in files to check whether unusual or unexpected characters are in any of the Physionet Challenge files.

efg | 2019-02-23

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
time.1 <- Sys.time()  
```

"Character Count" programs in this directory:

* charcnt2.exe - command line program for Linux of Windows.  Ouput vector of 256 character counts, along with total, in single line.

Note:  Source code for programs can be found on [GitHub](https://github.com/EarlGlynn/CharacterCount).

```{r}
library(readr)      # read_csv
library(kableExtra) # kable_styling
```

## Make list of .csv data files

```{r}
dataDIR <- "../../data/training" 
files <- list.files(dataDIR, full.names=TRUE)
```

## Get character counts for all files using charcnt2

Append each row of results to countsFile

```{r}
countsFile <- "##SaveIt.txt"
if (file.exists(countsFile)) file.remove(countsFile)

for (i in 1:length(files))
{
  if (i %% 500 == 0) cat(i, files[i], "\n") 
  shell(paste0(getwd(), "/charcnt2 ", files[i], " >> ", countsFile))
}
```

## Cleanup results

```{r}
results <- readLines(countsFile)
results <- substr(results, nchar(dataDIR)+2, nchar(results))
results <- c(paste0("Decimal,", paste(0:255, collapse=","), ",Total"),
             results)
file.remove(countsFile)
```

## Save results

```{r}
filename <- "character-counts.csv" 
writeLines(results, filename)
```

Use Excel to make the output more readable.

Normally edit the .csv file above (saving as an .xlsx file) to clarify that "Decimal" is the row label and that filenames are in the first column.  Add a TOTAL row for column sums.

Normally the Excel file is edited to show the first three rows as the Decimal, Hex and Character versions of the possible printable characters in a file.

```{r, echo=FALSE}
time.2 <- Sys.time()
processingTime <- paste("Processing time:", sprintf("%.1f",
                        as.numeric(difftime(time.2,
                                            time.1, units="secs"))), "secs\n")
```

`r processingTime`
`r format(time.2, "%Y-%m-%d %H%M")`                      